---
title: "Introduction to R: Graphs and starting with stats"
author: "David Pupovac"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_caption: yes
    fig_width: 6
    theme: readable
  pdf_document:
    fig_caption: yes
    fig_height: 3
    fig_width: 4
fontsize: 10pt
---
  
<style type="text/css">
  
  body, td {
    font-size: 14px;
    text-align: justify;
  }
code.r{
  font-size: 12px;
}
pre{
  background-color: #FFFFFF;
    font-size: 12px;
}
pre:not([class]) {
  background-color: #D8D8D8;
    color: black;
}
h1 {
  font-size: 28px;
  text-align: center;
}
h2 {
  font-size: 24px;
  text-align: left;
}
h3 {
  font-size: 18px;
  text-align: left;
  margin: 2em 0 1.5em 0;
}
h4 {
  font-size: 18px;
  text-align: center;
  margin: 0em 0 1em 0;
}
</style>
---  


```{R}
# setwd("C:/Desktop/Introduction to R")
```

```{R}
EUnoNA <- na.omit(read.csv("C:/Users/Korisnik.Korisnik-PC/Desktop/Lazarski/Introduction to R/third_class_data/example.csv")) # notice: we combined several commands together 
```
 
<br>

**Lets do some basic graphs first** 
 
<br>

###Histogram

```{R}
hist(EUnoNA$polar)
```

Let's make it a bit more fancy 
```{R}
hist(EUnoNA$polar, main="Histogram of polarization",
xlab = "Polarization", col ="steelblue")
```

To get two graphs in one
```{R}
layout(1:2)

hist(EUnoNA$polar, main="Histogram of polarization",
xlab = "Polarization", col ="darkred")

hist(EUnoNA$GDP, main="Histogram of GDP",
xlab = "GDP", col ="darkgreen")
```

**To save graphs click Plots (lower right window), click Export**
 
<br>

###Scatterplots
```{R}
layout(1:1) # fixing the layout 

plot(EUnoNA$polar,EUnoNA$nopar)
```

Let's make it a bit more fancy again

```{R}
plot(EUnoNA$polar,EUnoNA$nopar, main="Scatterplot of polarization vs. effective number of parties",
ylab = "Polarization", xlab = "Effective number of parties", col = "red", pch = 18)
```

What if you want to see scatterplots for every pair of variables?
```{R}

# install.packages("car")
library(car)

pairs(~polar + vol + nopart + disp + GDP +  demo + nogov, data=EUnoNA, 
      main="Simple Scatterplot Matrix")
```

Notice that in this case we did not refer to variable as "variable$dataset" but rather we specified a dataset with command data=EUnoNA and used names of variables (column headers).

###Boxplots

```{R}
boxplot(EUnoNA$polar, main="Boxplot of polarization", col="gray")
```


###ggplot2

R has powerful graphing packages that allow you to create more esthetically appealing graphs. The best known packages are ggplot2 and lattice. Let's just do one relatively simple example using ggplot2

```{r}
#install.packages("ggplot2")

library(ggplot2)

ggplot(EUnoNA, aes(y=EUnoNA$nopar, x=EUnoNA$polar)) + 
  geom_point() +
  geom_smooth(method=lm, color="black") +
  labs(title="Scatterplot with fit line - polarization vs. effective number of parties",
       x="Polarization", y = "Effective number of parties")
```  

There are numerous possibilities with these functions. Let us just try to change the theme.
  
```{r}  
# We can assign the graphing function to an object

p <- ggplot(EUnoNA, aes(y=EUnoNA$nopar, x=EUnoNA$polar)) + 
  geom_point() +
  geom_smooth(method=lm, color="black") +
  labs(title="Scatterplot with fit line - polarization vs. effective number of parties",
       x="Polarization", y = "Effective number of parties") 

# Notice that you only specify an additional element with +

p + theme_classic() 

p + theme_minimal()
```

Let us try to add some lines to the graph

```{r}

# Add a horizontal line at mean of y 
p + geom_hline(yintercept=mean(EUnoNA$nopar), linetype="dashed", color = "red", size=2) 

# Add a vertical line at mean of x
p + geom_vline(xintercept=mean(EUnoNA$polar), linetype="dotted", color = "darkgreen", size=1.5) 

# Notice the difference between geom_vline and geom_hline
```

###Standardization - z scores

But before we discus z scores we will discuss a powerful feature of most programming languages - **functions**. If you have to do something more than twice, write a function for it!  Here's a simple example. 

```{r}
standardize <- function(x) {
  (x - mean(x, na.rm = T))/sd(x, na.rm = T)
}  
# {} marks code chunk that is executed together. In the () we specify the arguments the function takes, in this case a vector that we name x
# You select the whole code chunk of the function, and run it.
```

x  -  notice, when looking in our global environment where we work, x is not defined. x is only defined within the local environment of the function. You can use various names
 
<br>

**Back to standardization:**

- Standardization assumes that a variable is **scaled** so that standard deviation of the new variable is 1
- Standardized variables can be compared to other standardized variables 
- The sign tells whether the score is located above (+) or below (-) the mean
- Standardization is directly related to the probability of obtaining certain outcomes by means of random selection from normal distributions. 

```{r}
sd(EUnoNA$nopart)
hist(EUnoNA$nopart, col="darkred", main="")

# So lets standardize it
EUnoNA$nopartStand <- standardize(EUnoNA$nopart)  ## we have created a new variable in our data that contains the standardized age
sd(EUnoNA$nopartStand)
hist(EUnoNA$nopartStand, col="darkgreen", main="")
```

Something is not right here! These variables are supposed to be identically distributed but on a different scale. How come they are not? Let's try to change the width of histogram bars

```{r}
hist(EUnoNA$nopartStand, col="gray38",  breaks = 8, main="")
```

This seems better, but still histograms are different. To definitely check, make a scatterplot of standardized and non-standardized variables. If everything is alright the scatterplot should be a straight line.

```{r}
plot(EUnoNA$nopart,EUnoNA$nopartStand)
```

<br>

###Central limit theorem

**Central limit theorem**: For any population with mean M and standard deviation SD, the distribution of sample means for sample size N will have a mean of M and a standard deviation of  SD/sqrt(N) and it will be approximately normally distributed. 

<br>

**Standard deviation** - standard distance between a score and the mean

**Standard error** - standard distance between sample mean and the population mean

<br>

So how do we get standard errors? The simplest way is using packages. 

```{r}
#install.packages("psych", dependencies = T)
library(psych)

(EUnoNADesc <- psych::describe(EUnoNA))  # Asterix marks non-numeric variables, so of course some statistics are misleading.
```

```{r}
str(EUnoNADesc)
```

```{r}
EUnoNADesc$se
```

Want the standard error of particular variables (e.g. effective number of parties)?  
```{r}
EUnoNADesc$se[4]
```

<br>

**Standard error** writing your own function 
```{r}
se <- function(x) sqrt(var(x,na.rm=TRUE)/length(na.omit(x))) #function for standard error

se(EUnoNA$nopart)
```
 
<br>

**Get confidence intervals**
```{r}
mean(EUnoNA$nopart, na.omit = T) # Check mean (I added to drop missing but we don't have any in this case)
```

**Lower CI 95%**
```{r}
mean(EUnoNA$nopart, na.omit = T) - se(EUnoNA$nopart) * qt(0.025, length(na.omit(EUnoNA$nopart)) - 1, lower.tail = F)
```

**Upper CI 95%**
```{r}
mean(EUnoNA$nopart, na.omit = T) + se(EUnoNA$nopart) * qt(0.025, length(na.omit(EUnoNA$nopart)) - 1, lower.tail = F)
```

Notice the qt command that gets us the critical t value for 0.025 probability in the tail. Degrees of freedom is the length of the thing that we calculate the CI for. Finally, -1 because degrees of freedom is n-1 for this calculation. However, often the researchers just use critical value (e.g. 1.96). 

<br>

###Chi-square test

Chi-square is used to assess if there is a relation between two categorical variables. First we need to recode variables

```{r}
# recode first variable

EUnoNA$stability <- "Low"

EUnoNA$stability[(EUnoNA$polar <= 5) & (EUnoNA$nogov <= 8)] <- "High"  
EUnoNA$stability[(EUnoNA$polar > 5 & EUnoNA$polar <= 8) & (EUnoNA$nogov > 8 & EUnoNA$nogov <= 11)] <- "Medium"
EUnoNA$stabFactor <- factor(EUnoNA$stability, levels = c("Low", "Medium", "High"), ordered = T)

# recode second variable

EUnoNA$wealth <- 1
EUnoNA$wealth[EUnoNA$GDP < median(EUnoNA$GDP)] <- 0 
```

There are couple of options to do a chi-square test of independence

**Option  1**

The package we want is called descr

```{r}
# install.packages("descr", dependencies = T)
library(descr)

crosstab(EUnoNA$wealth, EUnoNA$stabFactor, chisq = T, plot = F) #  the crosstab function does not create a useful plot
```

We get a warning message. Why do we get this? There must be at least 5 frequencies in each cell

**Option  2**

It is possible to do chi-square test without the cross-tab too
```{R}
chisq.test(EUnoNA$wealth, EUnoNA$stabFactor)
```


<br>

###t-tests

Lets look at the t.test command's help.

```{}
?t.test
```

You can set confidence level with conf.level and you can set if it is a paired sample or not. You can do two tailed (default) or one tailed (**never!!!**). 
 
<br>

#### One Sample t-test
Is the effective number of parties different than 3? 

```{r}
t.test(EUnoNA$nopart, mu=3)
```

<br>

####Independent Sample t test  

Let's check if the wealthier countries differ in terms of political polarization from the poorer ones

```{r}
(wealth_split <- split(EUnoNA$polar, EUnoNA$wealth))
x <- as.vector(wealth_split[[1]])
y <- as.vector(wealth_split[[2]])
t.test(x,y)
```
