---
title: 'Introduction to R: Essentials'
author: 'David Pupovac'
date: '`r Sys.Date()`'
output:
  html_document:
    fig_caption: yes
    fig_width: 6
    theme: readable
  pdf_document:
    fig_caption: yes
    fig_height: 3
    fig_width: 4
fontsize: 10pt
---
  
<style type="text/css">
  
  body, td {
    font-size: 14px;
    text-align: justify;
  }
code.r{
  font-size: 12px;
}
pre{
  background-color: #FFFFFF;
    font-size: 12px;
}
pre:not([class]) {
  background-color: #D8D8D8;
    color: black;
}
h1 {
  font-size: 28px;
  text-align: center;
}
h2 {
  font-size: 24px;
  text-align: left;
}
h3 {
  font-size: 12px;
  text-align: left;
  margin: 2em 0 1.5em 0;
}
h4 {
  font-size: 18px;
  text-align: center;
  margin: 0em 0 1em 0;
}
</style>
---  
<!-- you made a big font with #####--> 
#### What is R?

**The R is an environment for data analysis and graphics.**  

- Open source
- A full-featured programming language
- Two major parts: the base system and a collection of user contributed add-on packages

### Other Software

- [Phyton](https://www.python.org/)
- [PSPP](https://www.gnu.org/software/pspp/)   

### Pros and Cons

- Powerful
- Steep learning curve 

#### Information on R 

### Main web pages

Home site: [R-project](http://www.R-project.org)  
Comprehensive R Archive Network site: [CRAN](https://cran.r-project.org/)

### Internet (tremendous amount of information about R on the web). Some good sources:

- [Stack Overflow](http://stackoverflow.com/)
- [R bloggers](https://www.r-bloggers.com/)
- [Stack Exchange](stackexchange.com)

### Books on R - ever-growing number

- Springer series [Use R!](http://www.springer.com/series/6991)

### Some good 'beginner' books:

- Crawley, M J.  (2007) The R Book, John Wiley & Sons Ltd 
- Everitt, E. S., Hothorn (2009) A Handbook of Statistical Analyses using R, CRC Press
- Venables, W. N., Smith, D. M.  and the R Development Core Team (2016) An Introduction to R [get it here!](https://cran.r-project.org/doc/manuals/r-release/R-intro.pdf)

<br>

#### R and R Studio 

Available for mac, linux and windows

- [R](https://cran.r-project.org/mirrors.html)
- [R Studio](https://www.rstudio.com/products/rstudio/download/)

IDE (Integrated Development Environment/TextEditor). There are others.  emacs and vi and sublime text (even steeper learning curve). We use programing languages via IDEs.

### Download and install R and RStudio.  

With respect R you want the 'base' package and you want to run the setup program which will have a name like .exe (on a PC) or .dmg (on a Mac).

Manuals are available in R itself by choosing Help/Manuals (in PDF) from the menu bar.

The symbol > is prompt, it is R's way of saying 'What now?'.

You will sometimes see + instead of >. This means that the last command you typed is incomplete

<br>

#### Some general advice on scripting:

- **Never code in the console!** 
- **Always backup your data (i.e. use DropBox)**
- Never modify the original data - partly applicable to R (if you want to save data set with some changes, use another name)
- Running code ctrl+r (or ctrl+enter) on Windows, cmd+enter on mac 
- You can also use Run in RStudio (highlight the section or place the cursor somewhere on the line)
- In console, up and down arrow are useful
- If something goes wrong, check "small" things first: spelling; omitted comma, semicolon; uppercase/lowercase; brackets, data type...
- Press the Esc key if something goes wrong and the command line prompt > will reappear.
- Saving (not the console, the script) -  use .R at the end.
- Keep the script tidy: probably too early (but it is never too early), see here how to script well 
[clicky1](https://google.github.io/styleguide/Rguide.xml), [clicky2](http://adv-r.had.co.nz/Style.html)
- Be sure to comment your code. We insert comments in R using # 

So: 

```{r}
# write whatever here
```

does nothing. So, to make a separation in code you can do the following: 

```{r}
# =======================================================
#        Introduction to R: Essential stuff      
# =======================================================

# -------------------------------------------------------
# This demonstrates how to insert separations,comments and organize the code. Now we will run some real commands.
```


<br>

#### Running some commands from introductory message:
<!-- you forced new line by making space--> 
citation()  
help.start()  
demo()  
demo(graphics) 

####Basic commands (built-in functions):

```{r}
2 + 2
8 - 3
```
You can continue a complicated expression to the next line:
```{r}
5+6+3+6+4+2+4+8+
 3+2+7
```
Two or more expressions can be placed on a single line:
```{r}
2+3; 5*7; 3-7
```

### Arithmetic operators: 

- equals: =
- addition: +     
- subtraction:  -     
- multiplication: *     
- division: /     
- power: ^     

```{r}
2*(4+2)  
2*4+2  
8^(1/3)  
```

### Some mathematical functions

- sqrt(x)
- exp(x)
- log(x) 
- log10(x) 
- factorial(x)
- sin(x)
- abs(x)
- round(x, digits=0)

```{r}
sqrt(25) + 2
log10(1000)
abs(-45)
round(4.35689, digits=2)
```

### Logical arguments:

- not: !  
- or,union: |
- and,intersection: & 
- larger than: >
- equal: == 
- not equal: != 

```{r}
26 == (3 * 13) - 13
26 == (3 * 13) - 12
26 != (3 * 13) - 12
3 < 4
3 > 4
((4>8) | (4>3))
((4>8) & (4>3))
```

### Characters:

```{r}
print('Hello World')
```
You can use both double or simple quotes, just keep it consistent

<br>

#### Working directory

Get current working directory using getw:
```{r}
getwd()
```

- **R  uses forward slash or double backward slash**
- Use meaningful one, keep everyting organized!

Set new working directory using setwd()
```{r}
setwd('C:/Users/Korisnik.Korisnik-PC/Desktop/Lazarski/Introduction to R')
getwd()
```

<br>

#### Objects

R is [object-oriented programming language](https://en.wikipedia.org/wiki/Object-oriented_programming)

Objects obtain values in R by assignment. This is achieved by the gets arrow: <-

Be careful these two are not the same:
```{r}
y <- 10
y < -10 # this one is an error 
```
See the object
```{r}
y
print(y) # same but more elegant
```

Also works with logical statements

```{r}
x <- T
x
```

R is not going to stop you from running invalid calculations (but it still won't perform for example calculations with strings)

```{r}
# Another type of error: TRUE is interpreted as 1 and FALSE as 0
z <- x+y
z
```

**Beware of rewriting objects, you can't undo it!**

<br>

##### Main data types in R:

- **VECTORS**
- **MATIRCES**
- **ARRAYS**
- **LISTS**
- **DATAFRAMES**

#### Vectors

Vectors are variables with one or more values of the same type: logical, integer, real, complex, string (or character) or raw. 

### Creating vectors

Defined with  the c command, like this:
```{r}
numericVector <- c(1,2,3,4,5)
print(numericVector)
numericVector <- c(1:5)      # notice how sequence is defined 
print (numericVector)
numericVector <- c(1:5,0.63) # the numbers are no longer integers but floats   
print (numericVector)
```
You can enter values by hand, pressing the Enter key:

- numericVector <- scan()

You can copy paste from Excel with scan. In Excel, highlight the column of numbers you want, then type Ctrl+C (make sure the numbers are in column, not in row). Now go back into R. At the 1: prompt just type Ctrl+V.

```{r}
charvector <- c('Bob', 'Jane', 'Jack')
print (charvector)

logicalvector <- c(T, F, T)
print (logicalvector)
```
You cannot mix the values:
```{r}
MixingAttempt <- c(1, F, T)
print (MixingAttempt)
MixingAttempt <- c(1, 'Bob', -0.23, 'Jack')
print (MixingAttempt)
```

