---
title: "Introduction to R: Data Frames"
author: "David Pupovac"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_caption: yes
    fig_width: 6
    theme: readable
  pdf_document:
    fig_caption: yes
    fig_height: 3
    fig_width: 4
fontsize: 10pt
---
  
<style type="text/css">
  
  body, td {
    font-size: 14px;
    text-align: justify;
  }
code.r{
  font-size: 12px;
}
pre{
  background-color: #FFFFFF;
    font-size: 12px;
}
pre:not([class]) {
  background-color: #D8D8D8;
    color: black;
}
h1 {
  font-size: 28px;
  text-align: center;
}
h2 {
  font-size: 24px;
  text-align: left;
}
h3 {
  font-size: 16px;
  text-align: left;
  margin: 2em 0 1.5em 0;
}
h4 {
  font-size: 18px;
  text-align: center;
  margin: 0em 0 1em 0;
}
</style>
---  

#####Dataframes

A dataframe is an object with rows and columns (a bit like a matrix). **The values in the body of a dataframe can be numbers, but also be text.** A data frame is a list of equal-length vectors. 
 
<br>

## Dataframes and other data types

Create a list:
```{r}
(newList <- list(x = c(2, 5), y = c(6, 8.1)))
newList$x 
```

Convert list to a dataframe:
```{r}
(newData <- as.data.frame(newList))
```

Convert matrix to a dataframe (3 rows, 4 columns):

```{r}
(newData <- as.data.frame(matrix(1:12,3,4))) # notice the names at the top
```
 
<br>

### Subsetting dataframes

Let's give nice names to variables first

```{r}
colnames(newData) <- c("question.1","question.2","question.3","question.4")

# to see dataframe variable names:
colnames(newData)
```

Subsetting method is identical to the method used for matrices:  

```{r}
newData[2:3]   # select second and third column
newData[2,3]   # select an element in second row and third column
newData[2,]    # select second row
newData[,2]    # select second column
newData[c("question.1", "question.3")] # select using variable names
```

**However, most commonly to select a single variable from a dataframe you will use $ in form of "data$variable". See examples below:** 
```{r}
# First let's load some data which is already distributed with R 
dataIris <- data.frame(iris)
dataIris
# data$Species
```

Selecting variables:
```{r}
subset_iris <- dataIris$Sepal.Length
subset_iris
```

Data frames collect different types of vectors. We can subset the data frames on the basis of vector types
```{r}
subset_iris <- dataIris[,sapply(dataIris,is.numeric)] # select all numeric columns
subset_iris <- dataIris[,sapply(dataIris,is.factor)]  # select all factor columns
```

Selecting a subset of vectors:
```{r}
# to exclude variables: Sepal.Length, Sepal.Width, Species: 
myvars <- names(dataIris) %in% c("Sepal.Length", "Sepal.Width", "Species")
newdata <- dataIris[!myvars]

# to select variables: Sepal.Length, Sepal.Width, Species just drop ! sign:
newdata <- dataIris[myvars]
```

Selecting a subset of cases:
```{r}
subset_iris <- dataIris[dataIris$Species == "versicolor",]       # select on the basis of factor values
subset_iris <- dataIris[!(dataIris$Species == "versicolor"),]    # negative selection based on factor
subset_iris <- dataIris[-which(dataIris$Species=="versicolor"),] # the same as above

# select using several variables: 
subset_iris <- dataIris[dataIris$Sepal.Length > median(dataIris$Sepal.Length) & dataIris$Sepal.Width < 2.6,]
```
 
<br>

### Adding and deleting variables

Remember that we can also use cbind!

```{r}
dataIris$new.variable <- 1:150

# see what happens when you try to add a shorter vector 
# dataIris$new.variable <- 1:130

# to delete a variable 
dataIris$new.variable <- NULL 
```

<br>

### Ways to look at your data

```{r}
View(dataIris) # show you the data like in SPSS, Stata..., in new tab

head(dataIris)  ## display first 6 rows
tail(dataIris)  ## display last 6 rows

# Naturally, these are the same as: 
dataIris[1:6, ]  
dataIris[145:150, ]  

dim(dataIris)  ## again, know the dimensions! (That is what the dim commands does)

str(dataIris)  ## very useful to get an idea what you are up against, but also to check whether you have the data frame and format with the variables you wanted to have 

# If you really must edit your data from R (and in principle you should not edit your data) you can:
fix(dataIris) 
```
 
<br>

### Sorting dataframe
```{r}
dataIris <- dataIris[order(dataIris$Petal.Length),]
dataIris[1:15, ]

dataIris <- dataIris[rev(order(dataIris$Petal.Length)),]  # reverse order
dataIris[1:15, ]

dataIris <- dataIris[order(dataIris$Petal.Length, dataIris$Sepal.Width),] # two variables
dataIris[1:15, ]
```
 
<br>

### Naming rows

- You might want to use row names: so far, it was just a number
- You most certainly can do this. In the long run, though, use unique ID variables in your data
- Row names can be saved, but better to always keep a unique ID

Suppose your observations are individuals, and you want a simple row name: Individual and the number. Let us generate a character vector of length 3 to do this.

```{r}
newData

(rNames <- paste("Individual", 1:3, sep = "_"))  
```
or

```{r}
rNames <- paste("Individual", 1:length(newData$question.1), sep = "_")  # why is this better?
rownames(newData) <- rNames  # as you can see, assign our new character vector to the row names of our data. You can use return values from a function on both sides of an assignment
newData[1:3, ]

dim(newData)  # Note, length(data frame) is not the way to go. length is good for vectors

length(newData) # compare this to:
length(newData$question.1)
```

```{r}
summary(newData)

str(newData)
```
 
<br>

## Importing data

**Some basics about data storage in general (before loading the data):**

- All the values of the same variable must go in the same column.
- Any spaces in any of the variable names (row 1) should be deleted (so, do not name variable "variable 1")
- It is good idea to remove any apostrophes from variable names
- Be sure you data has a id variable (although you can add your id variable in R)
- When working with data, we need to make sure missing values are coded correctly.  More on this later.
 
<br>

### Let's start importing data saved in various formats 

- Set working directory

```{R}
setwd('C:/Desktop/Introduction to R) # or whatever you want to use
```

- Put the data in that folder.  (**All of them.**)

- Now we can start importing data of different types

### Importing files of type .txt 
```{R}
myData <- read.table("example.txt", header=TRUE, sep="\t") 

# if working directory is not set, you can directly refer to the data
myData <- read.table("C:/Introduction to R/example.txt", header=TRUE, sep="\t") 

# You can also select data
myData<-read.table(file.choose(), header = TRUE, sep="\t")
```
### Importing files of type .csv  
```{R}
myData <- read.csv("example.csv", header=TRUE) 
```

### Importing files of type .xlsx (Excel)
```{r}
## install.packages('XLConnect', dependencies = T)
library(XLConnect)
myData <- readWorksheet(loadWorkbook("C:/Introduction to R/example.txt/example.xlsx"),sheet=1)
```

### Importing files of type .sav (SPSS) 
```{R}
## install.packages('foreign', dependencies = T) 
library(foreign)

myData <- read.spss("example.sav")  ## very basic, and accordingly not working
is.list(myData)  ## again, this is not what you expect. It is a list.

is.data.frame(myData) ## but we want data frames

myData <- read.spss("example.sav", to.data.frame = T)  ## it is closer to what we want
summary(myData[8:10])  ## but those labels?! Maybe better without

myData <- read.spss("example.sav", to.data.frame = T, use.value.labels = F)  
```

### Importing files of type .dta (Stata) 

```{R}
myData <- read.dta("example.dta")  
is(myData)

# do not convert to factor
myData <- read.dta("example.dta", convert.factors = F)  
```
